The imprint of global magma oceans on exoplanet demographics

### 2022-12-20 src/tex/ms.tex
make mathematical notation consistent

\subsubsection{Synthetic planet populations with and without magma oceans}
\begin{note}
    Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
    To enable a quantitative assessment of this detectability, we injected the signal into the simulated planet sample before observing it with simulated surveys.
    While we search for the magma ocean pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
    Relevant properties are a planet's mass $M$, its instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
    We consider three different cases:
    \begin{itemize}
   \item[A.] No radius change.
    This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
    It also applies outside the runaway greenhouse limit, where we assume a solid planetary surface and no steam atmosphere.
        A planet that lies \textit{within} the runaway greenhouse limit but that is already completely desiccated will have the same radius. \todo{Mention if this case is considered in the model or not (as of now, no).}
   \item[B.] Radius inflation due to a steam atmosphere.
    We assigned planets within the runaway greenhouse limit steam atmospheres, which increase atmospheric scale heights significantly.
   \item[C.] As B, but including radius decrease of inflated planets due to water in the melt.
    Due to greenhouse forcing, planetary surfaces below steam atmospheres are thought to be molten.
    Part of the water will then partition into the melt, removing it from the steam atmosphere, which in turn decreases the radius inflation.
    This effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values.
    \end{itemize}
\end{note}

@@

\begin{note}
    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        h_{\mathrm{null}}(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The magma ocean hypothesis is then defined as
\begin{equation}\label{eq:mo_hypo}
    h_{\mathrm{MO}}(\theta, S) =
%        \begin{cases}
%        1 & \text{if } x \in \mathbb{Q}\\
%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
%    \end{cases}
        \begin{cases}
        h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
        <R_\mathrm{P}> (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
    \end{cases}
\end{equation}
    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $<R_\mathrm{P}>$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This magma ocean hypothesis can be tested agains the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
\end{note}

\begin{note}
    The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $<R_\mathrm{P}>$.
    The functional form of the magma ocean hypothesis is more complex: Besides the mean radius of planets outside the runaway greenhouse threshold $<R_\mathrm{P}>_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
    Further, the parameter encoding the amount of radius change is the planetary water mass fraction~$x_{H_2O}$.
\end{note}

@@

A sensible choice of priors is central for evidence estimation via nested sampling.
As the parameters we are interested in recovering, $S_\mathrm{thresh}$ and $x_{H_2O}$, are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
For $S_\mathrm{thresh}$, we chose a log-uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
Similarly, we sample $x_{H_2O}$ from a log-uniform distribution.
Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:mo_model}).
Finally, we adopt a broad, uniform prior for $<R_\mathrm{P}>$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
We then ran the nested sampler to compute the log-evidence and sample the posterior distributions.

>>>>

\subsubsection{Synthetic planet populations with and without magma oceans}
    Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
    To enable a quantitative assessment of this detectability, we injected the signal into the simulated planet sample before observing it with simulated surveys.
    While we search for the magma ocean pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
    Relevant properties are a planet's mass $M$, its instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
    We consider three different cases:
    \begin{itemize}
   \item[A.] No radius change.
    This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
    It also applies outside the runaway greenhouse limit, where we assume a solid planetary surface and no steam atmosphere.
        A planet that lies \textit{within} the runaway greenhouse limit but that is already completely desiccated will have the same radius. \todo{Mention if this case is considered in the model or not (as of now, no).}
   \item[B.] Radius inflation due to a steam atmosphere.
    We assigned planets within the runaway greenhouse limit steam atmospheres, which increase atmospheric scale heights significantly.
   \item[C.] As B, but including radius decrease of inflated planets due to water in the melt.
    Due to greenhouse forcing, planetary surfaces below steam atmospheres are thought to be molten.
    Part of the water will then partition into the melt, removing it from the steam atmosphere, which in turn decreases the radius inflation.
    This effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values.
    \end{itemize}

@@

    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        h_{\mathrm{null}}(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The magma ocean hypothesis is then defined as
\begin{equation}\label{eq:mo_hypo}
    h_{\mathrm{MO}}(\theta, S) =
%        \begin{cases}
%        1 & \text{if } x \in \mathbb{Q}\\
%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
%    \end{cases}
        \begin{cases}
        h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
        \langle R_\mathrm{P}\rangle (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
    \end{cases}
\end{equation}
    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This magma ocean hypothesis can be tested agains the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.

\begin{note}
    The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
    The functional form of the magma ocean hypothesis is more complex: Besides the mean radius of planets outside the runaway greenhouse threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
    Further, the parameter encoding the amount of radius change is the planetary water mass fraction~$x_{H_2O}$.
\end{note}

@@

A sensible choice of priors is central for evidence estimation via nested sampling.
As the parameters we are interested in recovering, $S_\mathrm{thresh}$ and $x_{H_2O}$, are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
For $S_\mathrm{thresh}$, we chose a log-uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
Similarly, we sample $x_{H_2O}$ from a log-uniform distribution.
Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:mo_model}).
Finally, we adopt a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
We then ran the nested sampler to compute the evidence and sample the posterior distributions.
